Rényi Entropy-Based Heterogeneity Detection in SAR Data

### 2025-03-15 Text/Articles/RENYI_ENTROPY_SAR_2025/R0-Renyi-entropy-sar.qmd
Update R0-Renyi-entropy-sar.qmd

Introduced by Alfréd Rényi in 1961\ [@renyi1961measures], this measure generalizes several well-known entropies, including the Shannon entropy\ [@Ribeiro2021]. For a continuous random variable $Z$ with pdf $f(z)$, the Rényi entropy of order $\lambda$, with $\lambda > 0$ and $\lambda \neq 1$, is defined as:
\begin{equation}
\label{E:entropy2}
H_\lambda(Z) = \frac{1}{1 - \lambda} \ln \int_{-\infty}^{\infty} [f(z)]^\lambda \, dz.
\end{equation}
Using \eqref{E:entropy2}, we derive closed-form expressions for the Rényi entropy of the $\Gamma_{\mathrm{SAR}}$ and $\mathcal{G}^0_I$ distributions:
\begin{multline}
\label{eq-HGammaSAR}
H_\lambda\bigl(\Gamma_{\text{SAR}}(L, \mu)\bigr)
= 
\ln \mu - \ln L + \frac{1}{1-\lambda}
\Bigl[
  -\lambda\,\ln\Gamma(L) \\  + \ln\Gamma\bigl(\lambda(L-1)+1\bigr)  - \bigl(\lambda(L-1)+1\bigr)\,\ln\lambda
\Bigr]
\end{multline}
and
\begin{multline}\label{eq-HGI0}
H_\lambda\bigl(\mathcal{G}^0_I(\mu, \alpha, L)\bigr)=\ln\mu-\ln L +\ln(-1 - \alpha)\\
+ \frac{1}{\,1 - \lambda\,}
\Bigl[
   \lambda\bigl(
      \ln\Gamma(L - \alpha)
      -\ln\Gamma(-\alpha)
      -\ln\Gamma(L)
   \bigr)\\
   +\ln\Gamma\bigl(\lambda(L - 1) + 1\bigr)
   +\ln\Gamma\bigl(\lambda(-\alpha + 1) - 1\bigr)\\
   -\ln\Gamma\bigl(\lambda(L - \alpha)\bigr)
\Bigr].
\end{multline}
The derivation of this result is given in the Appendix.

@fig-plot presents the Rényi entropy of $\mathcal{G}^0_I$ as a function of $\mu$ for different $\alpha$ values. As $\alpha \to -\infty$, it approaches the Rényi entropy of $\Gamma_{\text{SAR}}$, which is aligned with the fact that $\Gamma_{\text{SAR}}$ is a limiting case of the $\mathcal{G}^0_I$ model.

@@

Following Vasicek [@vasicek1976test] and Ebrahimi et al. [@Ebrahimi1994] for Shannon entropy estimation, and using the $m$-spacing density estimator, the Rényi entropy can be estimated as:
\begin{align}
\label{eq:est_R}
\widehat{H}_\lambda(\bm{Z}) = \frac{1}{1 - \lambda} \ln \left[\frac{1}{n} \sum_{i=1}^{n} \left( \frac{c_i m / n}{Z_{(i+m)} - Z_{(i-m)}} \right)^{\lambda - 1} \right].
\end{align}
This estimator is asymptotically consistent, i.e., it converges in
probability to the true value when $m,n\rightarrow\infty$ and
$m/n\rightarrow0$. 
We choose to use the heuristic formula for spacing, $m=\left[\sqrt{n}+0.5\right]$.

>>>>

Introduced by Alfréd Rényi in 1961\ [@renyi1961measures], this measure generalizes several well-known entropies, including the Shannon entropy\ [@Ribeiro2021]. For a continuous random variable $Z$ with pdf $f(z)$, the Rényi entropy of order $\lambda$, with $\lambda > 0$ and $\lambda \neq 1$, is defined as:
\begin{equation}
\label{E:entropy2}
H_\lambda(Z) = \frac{1}{1 - \lambda} \ln \int_{-\infty}^{\infty} [f(z)]^\lambda \, dz.
\end{equation}
Using \eqref{E:entropy2}, we derive closed-form expressions for the Rényi entropy of the $\Gamma_{\mathrm{SAR}}$ and $\mathcal{G}^0_I$ distributions:
\begin{multline}
\label{eq-HGammaSAR}
H_\lambda\bigl(\Gamma_{\text{SAR}}(L, \mu)\bigr)
= 
\ln \mu - \ln L + \frac{1}{1-\lambda}
\Bigl[
  -\lambda\,\ln\Gamma(L) \\  + \ln\Gamma\bigl(\lambda(L-1)+1\bigr)  - \bigl(\lambda(L-1)+1\bigr)\,\ln\lambda
\Bigr],
\end{multline}
and
\begin{multline}\label{eq-HGI0}
H_\lambda\bigl(\mathcal{G}^0_I(\mu, \alpha, L)\bigr)=\ln\mu-\ln L +\ln(-1 - \alpha)\\
+ \frac{1}{\,1 - \lambda\,}
\Bigl[
   \lambda\bigl(
      \ln\Gamma(L - \alpha)
      -\ln\Gamma(-\alpha)
      -\ln\Gamma(L)
   \bigr)\\
   +\ln\Gamma\bigl(\lambda(L - 1) + 1\bigr)
   +\ln\Gamma\bigl(\lambda(-\alpha + 1) - 1\bigr)\\
   -\ln\Gamma\bigl(\lambda(L - \alpha)\bigr)
\Bigr].
\end{multline}
These results are proven in the Appendix.

@fig-plot presents the Rényi entropy of the $\mathcal{G}^0_I$ model as a function of $\mu$ for different values of $\alpha$. 
As $\alpha \to -\infty$, it approaches the Rényi entropy of the $\Gamma_{\text{SAR}}$ law, in line with the fact that the latter is a limiting case of the former.

@@

Following Vasicek [@vasicek1976test] and Ebrahimi et al. [@Ebrahimi1994] for Shannon entropy estimation, and using the $m$-spacing density estimator, the Rényi entropy can be estimated as:
\begin{align}
\label{eq:est_R}
\widehat{H}_\lambda(\bm{Z}) = \frac{1}{1 - \lambda} \ln \left[\frac{1}{n} \sum_{i=1}^{n} \left( \frac{c_i m / n}{Z_{(i+m)} - Z_{(i-m)}} \right)^{\lambda - 1} \right].
\end{align}
This estimator is asymptotically consistent, i.e., it converges in
probability to the true value when $m,n\rightarrow\infty$ and
$m/n\rightarrow0$. 
We choose to use the heuristic spacing $m=\left[\sqrt{n}+0.5\right]$.
